Non-parametric approaches to optimal policy are more robust

### 2013-03-21 inst/doc/nonparametric-bayes.tex
updates to pomp during our discussion today

\begin{document}
\begin{frontmatter}
  \title{Non-parametric approaches to optimal policy are more robust}
  \author[cstar]{Carl Boettiger\corref{cor1}}
  \author[noaa]{Stephan B. Munch}
  \author[cstar]{Marc Mangel}
  \ead{cboettig@ucsc.edu}
  \cortext[cor1]{Corresponding author, cboettig@ucsc.edu}
  \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA 95064, USA}
  \address[noaa]{Southwest Fisheries Science Center, National Oceanic and Atmospheric Administration, 110 Shaffer Road, Santa Cruz, CA 95060, USA}
 \end{frontmatter}

@@

Process-based models built on modern machine learning approaches offer a
powerful and underappreciated approach for ecological forecasting and
management. We present a novel example of how one such example -
non-parametric Gaussian processes, can be adapted to the context of
management and decision-making.

While effective machine learning algorithms have speckled the
literature, most theoretical ecologists in particular have been wary of
these `black-box' approaches in favor of simple, mechanistic models.
While a blind application of machine learning approaches would indeed be
dangerous grounds to stake either ecological understanding or management
practice, such concerns are not grounds to reject such methods out of
hand. Rather, further attention is warranted to identify the
circumstances in which machine-learning approaches will most likely
outperform existing more mechanistic approaches, or help crack difficult
management questions where ecological or dynamical complexity has caused
traditional approaches to founder.

Here we illustrate how a machine learning model can be implemented in
place of simple mechanistic model in a decision theory (optimal control)
problem, and highlight the features of this system that make the
nonparametric approach particularly effective.

Mechanistic models have long been the gold standard of theoretical
modeling in ecology (e.g.~see @Geritz2012 or @Cuddington2013). Only by
understanding the processes involved can we make reliable long term
predictions and build an knowledge of cause and effect that guides the
hypotheses we make, the data we collect, and the management decisions we
make. Process-based models, whether expressed in the language of
mathematics or English, identify the connection between mosquitoes and
the spread of malaria, or greenhouses gases and climate change, guiding
our approach to understand and manage these threats. Despite this
central importance, we argue here that ecologists would do well to give
greater attention to the role non-mechanistic models can play in
ecological management and decision making. The value of these approaches
is greatest in a context where decisions are made over short time
horizons, and updated as new data becomes available.

\subsection{The alternative to process-based models: pattern-based
modeling}

Before we proceed further it would be useful to define our terms and lay
some greater context for the discussion. Throughout, we will distinguish
``mechanistic'' or ``process-based'' models from models that are merely
``correlative'' or ``pattern-based''. Historically pattern-based
modeling meant regression (usually linear regression) and resided in
methodology rooted in statistics departments and was the primary focus
of empirical ecologists, while mechanistic modeling meant dynamical
systems (ODEs, PDEs, SDEs and their discrete kin), residing in
methodology from the mathematics department and was the primary focus of
theoretical ecologists.

We believe these historical divides continue to color much of the
literature today, with theorists more skeptical than empiricists of
statistical methodology and vice versa. Meanwhile, the ground underneath
has shifted. Originally pattern-matching approaches could be critiqued
on the grounds of their simplicity (not everything is linear) while
mechanistic approaches could be critiqued on their tenuous connections
to data -- model parameter values such as death rates, birth rates, etc.
would be estimated in advance and then stuck into the model, rather than
estimated in the context of the model itself. Today both of these
critiques are outdated. Computational power and hierarchical statistical
methods (particularly approximate likelihood or simulation techniques
such as particle filters and approximate Bayesian computing) have
brought the inference richer dynamical systems into their fold, while
pattern-based modeling has spawned an entirely new approach under the
banner of machine learning that has divided the statistics community
(See @Bremman2001, ``The Two Cultures''). Machine learning models can
represent almost arbitrarily complex patterns and incorporate learning
and decision making strategies that have earned them the name
`artificial intelligence.' Yet at the same time they can be far less
transparent their linear regression predecessors, making both astounding
successes and startling biases and failures like their biological
namesake. More ecologists, particularly theorists and modelers, would do
well to learn both how to take advantage of these strengths and
recognize their weaknesses.

Mechanistic modeling emphasizes the importance of capturing the correct
gross properties of a system over tracking minute fluctuations. For
instance, in selecting and parameterizing model of a population of
conservation concern, we may be most interested in getting the long-term
behavior correct -- such as identifying if the dynamics support
persistence of the population -- rather than worrying how well they
reflect the year-to-year fluctuations. We would certainly have good
reason to prefer such a model over alternatives which are irreconcilable
to the most basic biological processes, such as unbounded growth, or
growth curves that do not pass through the origin in a closed system. So
it may come as a surprise to realize that such obviously wrong models
can perform as well or even better than reasonable mechanistic models in
guiding ecological management and decision making.

@@

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Multiple species * Online learning * Multiple step-ahead predictions *
  Explicitly accomidating additional uncertainties * Improving inference
  of optimal policy from the GP
\end{itemize}

\texttt{\{r echo=FALSE, results="asis"\} \#bibliography("html")}

>>>>

\begin{document}
\begin{frontmatter}
  \title{Non-parametric approaches to optimal policy are more robust}
  \author[cstar]{Carl Boettiger\corref{cor1}}
  \author[cstar]{Marc Mangel}
  \author[noaa]{Stephan B. Munch}
  \ead{cboettig@ucsc.edu}
  \cortext[cor1]{Corresponding author, cboettig@ucsc.edu}
  \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA 95064, USA}
  \address[noaa]{Southwest Fisheries Science Center, National Oceanic and Atmospheric Administration, 110 Shaffer Road, Santa Cruz, CA 95060, USA}
 \end{frontmatter}

@@

@@

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Multiple species
\item
  Online learning
\item
  Multiple step-ahead predictions
\item
  Explicitly accomidating additional uncertainties
\item
  Improving inference of optimal policy from the GP
\end{itemize}
